Tracing the Meta-Level: PyPy's Tracing JIT Compiler

### 2009-06-10 talk/icooolps2009/paper.tex
undo the smallification of the last commit.
Completely horrible hacks to squeeze everything on 8 pages again.

Typically tracing VMs go through various phases when they execute a program:
\begin{itemize}
\item Interpretation/profiling
\item Tracing
\item Code generation
\item Execution of the generated code
\end{itemize}

At first, when the program starts, everything is interpreted.
The interpreter does a small amount of lightweight profiling to establish which loops
are run most frequently. This lightweight profiling is usually done by having a counter on
each backward jump instruction that counts how often this particular backward jump
is executed. Since loops need a backward jump somewhere, this method looks for
loops in the user program.

When a hot loop is identified, the interpreter enters a special mode, called
\emph{tracing mode}. During tracing, the interpreter records a history of all
the operations it executes. It traces until it has recorded the execution of one
iteration of the hot loop. To decide when this is the case, the trace is
repeatedly checked during tracing as to whether the interpreter is at a position
in the program where it had been earlier.

The history recorded by the tracer is called a \emph{trace}: it is a sequential list of
operations, together with their actual operands and results. Such a trace can be
used to generate efficient machine code. This generated machine code is
immediately executable, and can be used in the next iteration of the loop.

@@

\begin{figure}
{
\begin{verbatim}
def f(a, b):
    if b % 46 == 41:
        return a - b
    else:
        return a + b
def strange_sum(n):
    result = 0
    while n >= 0:
        result = f(result, n)
        n -= 1
    return result

@@

Let's look at a small example. Take the (slightly contrived) RPython code in
Figure \ref{fig:simple-trace}.
The tracer interprets these functions in a bytecode format that is an encoding of
the intermediate representation of PyPy's translation tool\-chain after type
inference has been performed.
When the profiler discovers
that the \texttt{while} loop in \texttt{strange\_sum} is executed often the
tracing JIT will start to trace the execution of that loop.  The trace would
look as in the lower half of Figure \ref{fig:simple-trace}.

@@

The tracing JIT of the PyPy project is atypical in that it is not applied to the
user program, but to the interpreter running the user program. In this section
we will explore what problems this brings, and suggest how to solve them (at least
partially). This means that there are two interpreters involved, and we need appropriate
terminology to distinguish beween them. On the one hand, there is the interpreter that
the tracing JIT uses to perform tracing. This we will call the \emph{tracing
interpreter}. On the other hand, there is the interpreter that runs the
user's programs, which we will call the \emph{language interpreter}. In the
following, we will assume that the language interpreter is bytecode-based. The
program that the language interpreter executes we will call the \emph{user
program} (from the point of view of a VM author, the ``user'' is a programmer
using the VM).

Similarly, we need to distinguish loops at two different levels:
\emph{interpreter loops} are loops \emph{inside} the language interpreter. On
the other hand, \emph{user loops} are loops in the user program.

A tracing JIT compiler finds the hot loops of the program it is compiling. In
our case, this program is the language interpreter. The most important hot interpreter loop
is the bytecode dispatch loop (for many simple
interpreters it is also the only hot loop).  Tracing one iteration of this
loop means that
the recorded trace corresponds to execution of one opcode. This means that the
assumption made by the tracing JIT -- that several iterations of a hot loop
take the same or similar code paths -- is wrong in this case. It is very
unlikely that the same particular opcode is executed many times in a row.
\begin{figure}
\input{code/tlr-paper.py}
\caption{A very simple bytecode interpreter with registers and an accumulator.}
\label{fig:tlr-basic}
\end{figure}
\begin{figure}
{
\begin{verbatim}
    MOV_A_R     0   # i = a
    MOV_A_R     1   # copy of 'a'

@@

The condition for reusing already existing machine code also needs to be adapted to
this new situation. In a classical tracing JIT there is at most one piece of
assembler code per loop of the jitted program, which in our case is the language
interpreter. When applying the tracing JIT to the language interpreter as
described so far, \emph{all} pieces of assembler code correspond to the bytecode
dispatch loop of the language interpreter. However, they correspond to different
paths through the loop and different ways to unroll it. To ascertain which of them to use
when trying to enter assembler code again, the program counter of the language
interpreter needs to be checked. If it corresponds to the position key of one of
the pieces of assembler code, then this assembler code can be executed. This
check again only needs to be performed at the backward branches of the language
interpreter.

@@

Let's look at how hints would need to be applied to the example interpreter
from Figure \ref{fig:tlr-basic}. Figure \ref{fig:tlr-full} shows the relevant
parts of the interpreter with hints applied.  One needs to instantiate
\texttt{JitDriver} by listing all the variables of the bytecode loop.
The variables are classified into two groups, ``green'' variables and ``red''
variables. The green variables are those that the tracing JIT should consider to
be part of the program counter of the language interpreter. In the case of the
example, the \texttt{pc} variable is obviously part of the program counter;
however, the \texttt{bytecode} variable is also counted as green, since the
\texttt{pc} variable is meaningless without the knowledge of which bytecode
string is currently being interpreted. All other variables are red (the fact
that red variables need to be listed explicitly too is an implementation
detail).

@@

\begin{figure}
\input{code/no-green-folding.txt}
\caption{Trace when executing the Square function of Figure \ref{fig:square},
with the corresponding bytecodes as comments.}
\label{fig:trace-no-green-folding}
\end{figure}

@@

\begin{figure}[h]
\label{fig:bench-example}
{
\begin{verbatim}
def f(a):
    t = (1, 2, 3)
    i = 0
    while i < a:
        t = (t[1], t[2], t[0])
        i += t[0]
    return i
\end{verbatim}
}
\noindent
\begin{tabular}{|l|l|r|r|}
\hline
& &Time (s) &speedup\\
\hline
1 &PyPy compiled to C, no JIT &23.44 $\pm$ 0.07 &1.00\\
2 &PyPy comp'd to C, with JIT &3.58 $\pm$ 0.05 &6.54\\
3 &CPython 2.5.2 &4.96 $\pm$ 0.05 &4.73\\
4 &CPython 2.5.2 + Psyco 1.6 &1.51 $\pm$ 0.05 &15.57\\\hline
\end{tabular}

@@

There are quite a number of approaches that try to minimally enhance
interpreters to generate code at runtime without actually writing a native
compiler by hand. The goal of these is to get rid of dispatch overhead of
typical interpreters while retaining ease of implementation. Piumarta and
Riccardi \cite{piumarta_optimizing_1998} propose to copy fragments of the
interpreter together for commonly occurring bytecode sequences to reduce
dispatch overhead. However, dispatching is still needed to jump between such
sequences and also when non-copyable bytecodes occur. Ertl and Gregg
\cite{ertl_retargeting_2004} go further and stitch together the concatenated
sequences by patching the copied machine code. Thus they get rid of all dispatch
overhead. Both techniques can speed up interpreters which large dispatch
overhead a lot. However they will help less if the bytecodes themselves do a
lot of work (as is the case with Python
\cite{stefan_brunthaler_virtual-machine_2009}) and the dispatch overhead is lower. On
the other hand, our technique can do a better job by tracing inside the
implementation of those bytecode and inlining common paths.

@@

Furthermore both tracing and leaving machine code are very slow due to a
double interpretation overhead and we might need techniques for improving those.

>>>>

Typically tracing VMs go through various phases when they execute a program:
Interpretation/profiling, tracing, code generation and execution of the
generated code. When the program starts, everything is interpreted.
The interpreter does lightweight profiling to establish which loops
are run most frequently. This lightweight profiling is usually done by having a counter on
each backward jump instruction that counts how often this particular backward jump
is executed. Since loops need a backward jump somewhere, this method looks for
loops in the user program.

When a hot loop is identified, the interpreter enters a special mode, called
\emph{tracing mode}. During tracing, the interpreter records a history of all
the operations it executes. It traces until it has recorded the execution of one
iteration of the hot loop. To decide when this is the case, the trace is
repeatedly checked as to whether the interpreter is at a position
in the program where it had been earlier.

The history recorded by the tracer is called a \emph{trace}: it is a list of
operations, together with their actual operands and results. Such a trace can be
used to generate efficient machine code. This generated machine code is
immediately executable, and can be used in the next iteration of the loop.

@@

\begin{figure}
{\small
\begin{verbatim}
def f(a, b):
    if b % 46 == 41:
        return a - b
    else:
        return a + b
def strange_sum(n):
    result = 0
    while n >= 0:
        result = f(result, n)
        n -= 1
    return result

@@

As a small example, take the (slightly contrived) RPython code in
Figure \ref{fig:simple-trace}.
The tracer interprets these functions in a bytecode format that is an encoding of
the intermediate representation of PyPy's translation tool\-chain after type
inference has been performed.
When the profiler discovers
that the \texttt{while} loop in \texttt{strange\_sum} is executed often the
tracing JIT will start to trace the execution of that loop.  The trace would
look as in the lower half of Figure \ref{fig:simple-trace}.

@@

The tracing JIT of the PyPy project is atypical in that it is not applied to the
user program, but to the interpreter running the user program. In this section
we will explore what problems this brings, and suggest how to solve them (at least
partially). This means that there are two interpreters involved, and we need appropriate
terminology to distinguish beween them. On the one hand, there is the interpreter that
the tracing JIT uses to perform tracing. This we will call the \emph{tracing
interpreter}. On the other hand, there is the interpreter that runs the
user's programs, which we will call the \emph{language interpreter}. In the
following, we will assume that the language interpreter is bytecode-based. The
program that the language interpreter executes we will call the \emph{user
program} (from the point of view of a VM author, the ``user'' is a programmer
using the VM). Similarly, we need to distinguish loops at two different levels:
\emph{interpreter loops} are loops \emph{inside} the language interpreter. On
the other hand, \emph{user loops} are loops in the user program.

A tracing JIT compiler finds the hot loops of the program it is compiling. In
our case, this is the language interpreter. The most important hot interpreter loop
is the bytecode dispatch loop (for many simple
interpreters it is also the only hot loop).  One iteration of this loop
corresponds to the execution of one opcode. This means that the
assumption made by the tracing JIT -- that several iterations of a hot loop
take the same or similar code paths -- is wrong in this case. It is very
unlikely that the same particular opcode is executed many times in a row.
\begin{figure}
\input{code/tlr-paper.py}
\caption{A very simple bytecode interpreter with registers and an accumulator.}
\label{fig:tlr-basic}
\end{figure}
\begin{figure}
{\small
\begin{verbatim}
    MOV_A_R     0   # i = a
    MOV_A_R     1   # copy of 'a'

@@

The condition for reusing existing machine code also needs to be adapted to
this new situation. In a classical tracing JIT there is no or one piece of
assembler code per loop of the jitted program, which in our case is the language
interpreter. When applying the tracing JIT to the language interpreter as
described so far, \emph{all} pieces of assembler code correspond to the bytecode
dispatch loop of the language interpreter. However, they correspond to different
paths through the loop and different ways to unroll it. To ascertain which of them to use
when trying to enter assembler code again, the program counter of the language
interpreter needs to be checked. If it corresponds to the position key of one of
the pieces of assembler code, then this assembler code can be executed. This
check again only needs to be performed at the backward branches of the language
interpreter.

@@

Let's look at how hints would need to be applied to the example interpreter
from Figure \ref{fig:tlr-basic}. Figure \ref{fig:tlr-full} shows the relevant
parts of the interpreter with hints applied.  One needs to instantiate
\texttt{JitDriver} by listing all the variables of the bytecode loop.
The variables are classified into two groups, ``green'' variables and ``red''
variables. The green variables are those that the tracing JIT should consider to
be part of the program counter of the language interpreter. In the case of the
example, the \texttt{pc} variable is obviously part of the program counter;
however, the \texttt{bytecode} variable is also counted as green, since the
\texttt{pc} variable is meaningless without the knowledge of which bytecode
string is currently being interpreted. All other variables are red.

@@

\begin{figure}[t]
\input{code/no-green-folding.txt}
\caption{Trace when executing the Square function of Figure \ref{fig:square},
with the corresponding bytecodes as comments.}
\label{fig:trace-no-green-folding}
\end{figure}

@@

\begin{figure}[h]
\label{fig:bench-example}
{\small
\begin{verbatim}
def f(a):
    t = (1, 2, 3)
    i = 0
    while i < a:
        t = (t[1], t[2], t[0])
        i += t[0]
    return i
\end{verbatim}
}
\noindent
\begin{tabular}{|l|l|r|r|}
\hline
& &Time (s) &speedup\\
\hline
1 &PyPy compiled to C, no JIT &23.44 $\pm$ 0.07 &1.00\\
2 &PyPy comp'd to C, with JIT &3.58 $\pm$ 0.05 &6.54\\
3 &CPython 2.5.2 &4.96 $\pm$ 0.05 &4.73\\
4 &CPython 2.5.2 + Psyco 1.6 &1.51 $\pm$ 0.05 &15.57\\\hline
\end{tabular}

@@

There are quite a number of approaches that try to minimally enhance
interpreters to generate code at runtime without actually writing a native
compiler by hand. The goal of these is to get rid of dispatch overhead of
typical interpreters while retaining ease of implementation. Piumarta and
Riccardi \cite{piumarta_optimizing_1998} propose to copy fragments of the
interpreter together for commonly occurring bytecode sequences to reduce
dispatch overhead. However, dispatching is still needed to jump between such
sequences and also when non-copyable bytecodes occur. Ertl and Gregg
\cite{ertl_retargeting_2004} go further and get rid of all dispatch overhead by
stitching together the concatenated
sequences by patching the copied machine code. 
Both techniques can speed up interpreters which large dispatch
overhead a lot. However they will help less if the bytecodes themselves do a
lot of work (as is the case with Python
\cite{stefan_brunthaler_virtual-machine_2009}) and the dispatch overhead is lower. On
the other hand, our technique can do a better job by tracing inside the
implementation of those bytecode and inlining common paths.

@@
